Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-14 Manuscript/manuscript.Rmd
manuscript edits

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* the set of annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation on the distances (Supplementary Figures 1-2 E-H).

@@

We compared the performances of random forest classification models implemented on TAD boundaries called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines *L* = { *GM12878*, *K562* }, at four different resolutions *R* = { *10 kb*, *25 kb*, *50 kb*, *100 kb* }. For each classifier, we considered 3 separate predictor types as the feature space including *P* = {*OC*, *OP*, *Distance* }. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including *B* = {*None*, *RUS*, *ROS*, *SMOTE* }. Each combination of the above parameters was repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), *C* = {*1*, *2*, ..., *22*}. A system produced from the ensemble framework was defined as: *S* = {*c*, *r*, *p*, *b*}, where *l* $\in$ *L*, *r* $\in$ *R*, *p* $\in$ *P*, *b* $\in$ *B*, and *c* $\in$ *C*. Letting *|X|* denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2016 (= 2 $\times$ 4 $\times$ 3 $\times$ 4 $\times$ 21) separate models.

## IMBALANCE RATIOS REMAINED SEVERE AND CONSISTENT ACROSS RESOLUTIONS

Results showed that the imbalance ratios were well below acceptable thresholds for both cell lines, and across the 4 resolutions (Figure 4). The low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not. Likewise, there was very little variation in imbalance ratios across resolution. The low IR indicate a necessary re-sampling technique that will suitably balance the majority and minority classes.

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 3). Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases.

>>>>

To investigate the relationship between functional genomic annotations and the development of TAD boundaries, we devised 3 different predictor-types that composed the feature space of downstream random forest classification models. These predictor-type specific annotations represented the data matrix, **X** = {$X_{1}$, $X_{2}$, ..., $X_{p}$}, where *p* the set of annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C).  When measured in basepairs, distances were found to be highly skewed across all resolutions, for both the GM12878 and K562 cell lines (Supplementary Figures 1-2 A-D). As a normalization procedure, we performed a $log_{2}$ transformation, which resulted in much more gaussian-shaped distance densities (Supplementary Figures 1-2 E-H).

@@

We then developed an ensemble framework to systematically compare and establish the best set of data level characteristics for predicting TAD boundaries using random forest classifiers built on functional genomic annotations (Figure 3). TAD boundaries were called using the arrowhead algorithm [@durand2016juicebox] from replicate Hi-C data for 2 cell lines *L* = {*GM12878*, *K562*}, at four different resolutions *R* = {*10 kb*, *25 kb*, *50 kb*, *100 kb*}. For each classifier, we considered 3 separate predictor types as the feature space including *P* = {*OC*, *OP*, *Distance*}. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including *B* = {*None*, *RUS*, *ROS*, *SMOTE*}. Each combination of the above parameters was repeated 21 times (number of autosomal chromosomes, excluding chromosome 9), *C* = {*1*, *2*, ..., *22*}. A system produced from the ensemble framework was defined as: *S* = {*c*, *r*, *p*, *b*}, where *l* $\in$ *L*, *r* $\in$ *R*, *p* $\in$ *P*, *b* $\in$ *B*, and *c* $\in$ *C*. Letting *|X|* denote the cardinality of each set, we evaluated *|L|* $\times$ *|R|* $\times$ *|P|* $\times$ *|B|* different ensemble systems, totaling 2016 (= 2 $\times$ 4 $\times$ 3 $\times$ 4 $\times$ 21) separate models. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9; see Methods).

## IMBALANCE RATIOS REMAINED SEVERE AND INVARIANT TOWARD RESOLUTION

[??? include icc for each component of table 1 with resolution]
The total number of genomic bins, which defined the dimensions of the data for modelling, was contingent upon the resolution of the contact matrix that cell line specific TADs were called on. Higher resolutions resulted in smaller unit bin widths, which corresponded to larger numbers of genomic bins (Table 1). For GM12878, there was an average of 12,214.1, 4,892.7, 2,450.9, and 1,227.5 bins created across chromosomes, for each resolution respectively. Likewise, the average number of bins, respective to resolution, were found to be similar on the K562 cell line (Supplementary Table 3).

As expected, the number of TADs called was related to resolution. Higher resolutions were found to result in a greater number of TADs called. For GM12878, an average of 365.5, 222.4, 111.9, and 49.1 TADs were called across chromosomes, for each respective resolution (Table 1). Average TADs called across chromosomes were found to be less on the K562 cell line at all resolutions (Supplementary Table 3). Although the downward trend was still present. As a result, a greater number of TADs corresponded to a greater number of TAD boundaries. For GM12878, the average number of TAD boundaries across chromosomes was found to be 1,254.6, 714.0, 364.4, and 165.3 for each resolution respectively (Table 1). Similar boundary profiles were exhibited on the K562 cell line (Supplementary Table 3).

Furthermore, results showed that average imbalance ratios were consistently and severely low across resolutions, for both cell lines. For GM12878, the average imbalance ratios were found to be 0.12, 0.17, 0.17, and 0.15 for the 10 kb, 25 kb, 50 kb, and 100 kb resolutions respectively (Table 1). Slightly smaller resolution-specific and averaged imbalance ratios were featured on the K562 cell line, yet were still consistent and invariant to resolution. Such low imbalance ratios indicated a wide disparity between the number of genomic bins that overlaped with a flanked TAD boundary and the much larger number that did not.

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases. <!--Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 4).-->
